preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-07 Manuscript/manuscript.Rmd
Last part of results edits

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological;@krefting2018evolutionary]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) [??? Were they threated separately?] produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by PEAKACHU at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2).

@@

## Comparing called vs predicted boundaries

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? Is there other resolution? ??? TAD and loop boundaries should be distinguished, to describe when we have all results] [??? Define that ARROWHEAD data constitute the main results]

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? DEfine all indexes, what is N, p+1, i?]

<!-- Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for PEAKACHU defined boundaries (Supplementary Tables 3-4). -->

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions. These results were consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and loop boundary prediction.

@@

![**Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and PEAKACHU respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and PEAKACHU loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?]

@@

Subsequent comparisons of *preciseTAD* with both ARROWHEAD and PEAKACHU revealed that our predicted boundaries were more conserved between the GM12878 and K562 cell lines. To assess conservation, we evaluated the overlap between cell line specific boundaries after flanking by the resolution used to make predictions. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving PEAKACHU were flanked by 10 kb. We found that a significant proportion of flanked boundaries predicted by *preciseTAD* overlapped between GM12878 and K562. Specifically, 45%??? and 61%??? of *preciseTAD* boundaries were conserved across cell lines when built on ARROWHEAD and PEAKACHU boundaries respectively (Figure 6A). Additionally, the median jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and PEAKACHU boundaries were conserved (Figure 6A, Supplementary Table 5).

Expectedly, the majority of flanked TAD boundaries predicted by *preciseTAD* represented a subset of the larger group of predicted loop boundaries. This makes sense as the boundaries that demarcate chromatin loops make up a broader class of boundaries, with TAD boundaries representing the next level in 3D chromatin hierarchy, followed by A/B compartments. For GM12878, 83% of predicted TAD and loop boundaries overlapped with eachother, while only 3% of predicted TAD boundaries were not contained in the set of flank-predicted loop boundaries (Figure 6B). Similarly, for K562, 75% overlap occured between predicted TAD and loop boundaries, with only 4% of predicted TAD boundaries remaining outside of the set of flank-predicted loop boundaries (Figure 6B).

![**preciseTAD boundaries are conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the percentage of overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the nested structure of predicted TAD and loop boundaries identified by *preciseTAD* for the GM12878 and K562 cell lines. Note, predicted and called TAD boundaries were flanked by 5 kb, while predicted and called loop boundaries were flanked by 10 kb prior to calculating overlaps to reflect the resolution that was used to predict each respectively.](Figures/figure6_venn_diagrams.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the predictor space. We first optimized our RF model by creating a machine learning framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary region prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ genomic distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines.

@@

The performance of *preciseTAD* was found to be empirically better than both ARROWHEAD and PEAKACHU ground truth boundaries. *preciseTAD* provides bp resolution predicted boundaries, thereby alleviating resolution limitations of conventional TAD-callers. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* boundaries were found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance.

@@

**Supplementary Table 1.** A complete list of the Hi-C data libraries obtained from GEO GSE63525 [@rao20143d] used to obtain the locations of topologically associating domains for ARROWHEAD, as well as the locations for the cell line specific .txt files of loop boundaries obtained by PEAKACHU. [@salameh2019supervised].

@@

**Supplementary Table 4.** Loop boundary and class imbalance summaries from PEAKACHU 10 kb resolution for both the GM12878 and K562 cell lines respectively.

@@

**Supplementary Figure 4. Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.

**Supplementary Figure 8. preciseTAD boundaries are more enriched for GM12878-specific SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) PEAKACHU boundaries for the GM12878 cell line.

**Supplementary Figure 9.preciseTAD boundaries are more enriched for K562-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) PEAKACHU boundaries for the K562 cell line.

**Supplementary Figure 10. preciseTAD boundaries are spatially closer to GM12878-specific SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or PEAKACHU; blue) to the nearest SMC3, RAD21, and ZNF143 sites on the linear genome for GM12878.

**Supplementary Figure 11. preciseTAD boundaries are spatially closer to K572-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or PEAKACHU; blue) to the nearest CTCF, SMC3, RAD21, and ZNF143 sites on the linear genome for K562.

**Supplementary Table 5.** Median chromosome-specific Jaccard indices (IQR) for called and *preciseTAD* predicted TAD (ARROWHEAD) and loop (PEAKACHU) boundaries between the GM12878 and K562 cell lines. The p-value is from a nonparametric Wilcoxon-rank sum test. Note, predicted and called *TAD* boundaries were flanked by 5 kb, while predicted and called *loop* boundaries were flanked by 10 kb prior to calculating Jaccard indices to reflect the resolution that was used to predict each respectively.

@@

**Supplementary Table 7.** Additional performance metrics when implementing a random forest using PEAKACHU ground truth chromatin loop boundaries.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) [??? Were they threated separately?] produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by Peakachu at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2).

@@

## Comparing called vs. predicted boundaries

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? Is there other resolution? ??? TAD and loop boundaries should be distinguished, to describe when we have all results] [??? Define that ARROWHEAD data constitute the main results]

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? DEfine all indices, what is N, p+1, i?]

<!-- Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for Peakachu defined boundaries (Supplementary Tables 3-4). -->

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions. These results were consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on Peakachu loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and loop boundary prediction.

@@

![**Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and Peakachu respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?]

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell-line-specific boundaries detected by *preciseTAD*, ARROWHEAD, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that a significant proportion of flanked boundaries predicted by *preciseTAD* overlapped between GM12878 and K562. Specifically, 45%??? and 61%??? of *preciseTAD* boundaries were conserved across cell lines when built on ARROWHEAD and Peakachu boundaries respectively (Figure 6A). Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.

The majority of boundaries predicted by the ARROWHEAD-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Figure 6B). This is expected as loop boundaries (Peackachu) are more abundant, while TAD boundaries (ARROWHEAD) represent the narrower level of the 3D chromatin organization. 83% and 75% of predicted TAD and loop boundaries overlapped with each other for GM12878 and K562 cell lines, respectively (Figure 6B). Together with the results shown on Figure 4, the high overlap between boundaries predicted by ARROWHEAD- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?]

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the predictor space. We first optimized our RF model by creating a machine learning framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary region prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ genomic distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines.

@@

The performance of *preciseTAD* was found to be empirically better than both ARROWHEAD and Peakachu ground truth boundaries. *preciseTAD* provides bp resolution predicted boundaries, thereby alleviating resolution limitations of conventional TAD-callers. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* boundaries were found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance.

@@

**Supplementary Table 1.** A complete list of the Hi-C data libraries obtained from GEO GSE63525 [@rao20143d] used to obtain the locations of topologically associating domains for ARROWHEAD, as well as the locations for the cell line specific .txt files of loop boundaries obtained by Peakachu. [@salameh2019supervised].

@@

**Supplementary Table 4.** Loop boundary and class imbalance summaries from Peakachu 10 kb resolution for both the GM12878 and K562 cell lines respectively.

@@

**Supplementary Figure 4. Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.

**Supplementary Figure 8. preciseTAD boundaries are more enriched for GM12878-specific SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) Peakachu boundaries for the GM12878 cell line.

**Supplementary Figure 9.preciseTAD boundaries are more enriched for K562-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) Peakachu boundaries for the K562 cell line.

**Supplementary Figure 10. preciseTAD boundaries are spatially closer to GM12878-specific SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or Peakachu; blue) to the nearest SMC3, RAD21, and ZNF143 sites on the linear genome for GM12878.

**Supplementary Figure 11. preciseTAD boundaries are spatially closer to K572-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or Peakachu; blue) to the nearest CTCF, SMC3, RAD21, and ZNF143 sites on the linear genome for K562.
[??? Combine Supplementary Figure 10 and 11. Make panel A and, underneath, panel B. Remove CTCF from Y-axis and put overhead labels above each plot pair]

**Supplementary Table 5.** Median chromosome-specific Jaccard indices (IQR) for ARROWHEAD- and Peakachu-called and *preciseTAD* predicted boundaries between the GM12878 and K562 cell lines [??? Where is cell line information?] [??? What is in parentheses?]. The p-value is from a nonparametric Wilcoxon-rank sum test [??? Absent]. Note, ARROWHEAD-called and predicted boundaries were flanked by 5 kb, while Peakachu-called and predicted boundaries were flanked by 10 kb prior to calculating Jaccard indices.

@@

**Supplementary Table 7.** Additional performance metrics when implementing a random forest using Peakachu ground truth chromatin loop boundaries.
